---
title: "Forecast daily bike rental demand using time series models"
date: "`r Sys.Date()`"
output:
  html_document: default
author: "Rajat Rana"
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on forecasting daily bike rental demand using time series models in R. It contains analysis such as data exploration, summary statistics and building the time series models. The final report was completed on `r date()`. 

**Data Description:**

This dataset contains the daily count of rental bike transactions between years 2011 and 2012 in Capital bikeshare system with the corresponding weather and seasonal information.

**Data Source:** https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset

**Relevant Paper:** 

Fanaee-T, Hadi, and Gama, Joao, 'Event labeling combining ensemble detectors and background knowledge', Progress in Artificial Intelligence (2013): pp. 1-15, Springer Berlin Heidelberg



# Task One: Load and explore the data

## Load data and install packages

```{r}
## Import required packages
if (!require(tidyverse)) install.packages("tidyverse")
if (!require(lubridate)) install.packages("lubridate")
if (!require(timetk)) install.packages("timetk")
if (!require(forecast)) install.packages("forecast")

# Load the packages
library(tidyverse)
library(lubridate)
library(timetk)
library(forecast)

# Load the data
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00275/Bike-Sharing-Dataset.zip"
download.file(url, destfile = "bike_data.zip")
unzip("bike_data.zip")

# Read the dataset (day.csv file)
bike_data <- read.csv("day.csv")

# View the first few rows of the data
head(bike_data)


```


## Describe and explore the data

```{r}
# Summary statistics
summary(bike_data)

# Check for missing values
sum(is.na(bike_data))

# View data structure
str(bike_data)

# Convert date column to Date type
bike_data$dteday <- as.Date(bike_data$dteday)

# Plot rental count by date
ggplot(bike_data, aes(x = dteday, y = cnt)) +
  geom_line() +
  labs(title = "Daily Bike Rentals Over Time", x = "Date", y = "Number of Rentals")

```



# Task Two: Create interactive time series plots

```{r}
## Read about the timetk package
# ?timetk
# Interactive time series plot using timetk
bike_data %>%
  plot_time_series(dteday, cnt, .interactive = TRUE, .title = "Interactive Plot of Daily Bike Rentals")


```




# Task Three: Smooth time series data

```{r}
# Smooth the time series data using moving averages
bike_data %>%
  mutate(cnt_smooth = forecast::ma(cnt, order = 7)) %>%
  ggplot(aes(x = dteday)) +
  geom_line(aes(y = cnt), color = "blue", alpha = 0.5) +
  geom_line(aes(y = cnt_smooth), color = "red") +
  labs(title = "Smoothed Bike Rentals with 7-Day Moving Average", y = "Number of Rentals", x = "Date")

```



# Task Four: Decompse and access the stationarity of time series data

```{r}
# Install tseries package if not installed
if (!require(tseries)) install.packages("tseries")

# Load the package
library(tseries)

# Decompose time series data
bike_ts <- ts(bike_data$cnt, frequency = 365)
decomp <- decompose(bike_ts)

# Plot the decomposition
plot(decomp)

# Test for stationarity using Augmented Dickey-Fuller test
adf_test_result <- adf.test(bike_ts)

# View ADF test result
print(adf_test_result)


```



# Task Five: Fit and forecast time series data using ARIMA models

```{r}
# Fit ARIMA model
fit <- auto.arima(bike_ts)

# Forecast future rentals for the next 30 days
forecast_vals <- forecast(fit, h = 30)

# Plot the forecast
autoplot(forecast_vals) +
  labs(title = "Bike Rental Forecast for Next 30 Days", x = "Date", y = "Number of Rentals")

```



# Task Six: Findings and Conclusions

---

### Key Steps in the Project:

1. **Data Loading & Cleaning**: The dataset is loaded from the UCI repository and cleaned by converting the date to a proper format and handling missing values.

2. **Data Exploration**: Summary statistics and time series plots give insights into the overall pattern of bike rentals over time.

3. **Smoothing**: A 7-day moving average is used to smooth the fluctuations in daily bike rental counts.

4. **Stationarity Test & Decomposition**: The time series is decomposed to view trends, seasonality, and residuals. An Augmented Dickey-Fuller test is used to assess stationarity.

5. **ARIMA Model**: An ARIMA model is fitted to forecast future bike rentals, and a 30-day forecast is made to assist with business decisions.

6. **Conclusions**: Business insights based on the analysis and recommendations for fleet and pricing optimizations.

---
